Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges

### 2020-06-10 man/ecosystems/ms_ecosystems.Rmd
add statisticall methods

In the present study, we assessed resilience of *Quercus pyrenaica* Willd. (Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdoetal2013GrowthProjectionsa; @GeaIzquierdoetal2017RiskyFuture]. We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth). The analysis of resilience to climate was discussed taken into account the land-use history of these transformed forests. To do so we conducted a review of historical documents to reconstruct forest history in our study sites, and to determine how anthropogenic drivers have shaped the current forest structure. Based on this discussion of forest resilience, we specifically developed a rationale that integrates the ecological and anthropogenic dimensions of marginality to determine the regional and local scale mechanisms shaping the probability of persistence (or extinction) of oak rear edge populations. Our main hypothesis is that these stands will show low resilience to extreme droughts along a small-scale gradient. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their present geographical rear edge; (*ii*) conducted and explored a review of historical documents to reconstruct forest management history and to infer how it impacted tree-growth in time; (*iii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iv*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter, we characterized variability in the forest response to drought within the current geographical rear-edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.

@@

The forests of this species reach their southernmost European limit in Andalusian mountains such as Sierra Nevada (37°N, 3°W), a high-mountain range with elevations of up to 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability according to elevation. Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olalde2002]. Eight Pyrenean oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys. Today, *Quercus pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005]. They are the richest forest formation in vascular plant species of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008ABG]. They also harbor high levels of intraspecific genetic diversity [@Valbuena2013]. These relict forests have undergone intensive human use throughout history [@CamachoOlmedo2002]. Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013].

@@

### Field sampling and dendroecological methods to assess individual tree resilience
Trees were sampled during the autumn of 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1). For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit (CA-Low) and the tree-line (CA-High), respectively, in the site sampled. Despite the proximity of these two elevations (less than a 200-m difference) the stands differ markedly in their structure and characteristics (Table 1). The three sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015]. At each site, between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly sampled. Two cores of 5 mm in diameter were taken from each tree at breast height (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species, and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot with a 10-m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details].

@@

The growth trends were analyzed at different time scales. To study the growth response to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. These were calculated from ratios between raw growth measurements and individual cubic splines with a 50% frequency cutoff at 30 years [@Fritts1976]. Tree-ring width series were standardized and detrended using `dplR` [@Bunn2010]. Mean residual site chronologies were established by computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976]. The statistical quality of each chronology was checked via the expressed population signal (EPS). A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.

The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$). In theory, BAI represents a more accurate indicator of growth than ring width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile growth [@Biondi2008]. Raw ring widths and measured DBH were used to compute BAI [@Piovesan2008] with the following equation: $\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$ where $r$ is the radius of the tree and $t$ is the year of tree-ring formation. For each individual tree, a mean BAI series was calculated. Then, mean site BAI chronologies were determined by averaging individual tree BAI time series.

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (suppressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought-induced neighbor mortality) in the past. Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median [@Rubino2004].
```{r, eval=FALSE, echo=FALSE}
# Medians were used since they are a more robust estimator of central tendency than means [@Rubino2004].
```
Site-disturbance chronologies were constructed by annually averaging the individual disturbance series. To separate growth peaks caused by disturbance events and expressing stand-wise disturbances from those caused by climate, we considered a threshold of 50% of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014]. In addition, the history of the forest and management of our sampling sites was inferred from a detailed analysis of historical land-use changes. For this, existing historical documents were exhaustively reviewed to compile information on socio-economical activities affecting the forests being studied (Table S4). We reviewed several documentary sources: historical documents and maps; detailed mining reports; official information on recent wildfires events and forest-management practices; livestock farming; traditional irrigation channels; and studies concerning the socioeconomic dynamics of forests on Sierra Nevada at different scales (see Table S4 for references).

@@

### Statistical analysis 
The severe drought events since 1901 were identified using SPEI-12 and regional climatic data (Figure 2). They were characterized in terms of duration, severity, intensity [see @Spinoni2015]. In a first step we explored temporal trends of EVI and BAI variables. Temporal trends of $EVI_{mean}$ were examined at the pixel scale, using the Mann–Kendall nonparametric test. Temporal trends for BAI were assessed using mean BAI chronologies. The impact of drought in greenness and growth was explored using the $EVI_{sa}$ and the mean RWI site chronologies. Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim` [@Zang2015]. The non-climatic disturbance impacts on tree-growth were evaluated using the site disturbance chronologies.

@@

For each of the three resilience indices studied, we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the oak populations studied (northern and southern exposures). These tests were used because original and log-transformed data did not follow the assumptions of normality or homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used because they were close to the mean value in all cases [@Wilcox2012]. When the robust ANOVA test was run, data were bootstrapped 3000 times and trimmed automatically to control the potential influence of outliers. *Post-hoc* differences were assessed pairwise using a similar bootstrap test. All the robust ANOVA and *post-hoc* tests were carried out using the `WRS2` package. The level of significance was set to 0.05 and adjusted for multiple comparisons. Resilience metrics of BAI were also computed for the most severe drought events since 1950, and the relationship of them to the severity of the drought were explored.

>>>>

In the present study, we assessed resilience of *Quercus pyrenaica* Willd. (Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdo2013; @GeaIzquierdoetal2017RiskyFuture]. We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth). The analysis of resilience to climate was discussed taken into account the land-use history of these transformed forests. To do so we conducted a review of historical documents to reconstruct forest history in our study sites, and to determine how anthropogenic drivers have shaped the current forest structure. Based on this discussion of forest resilience, we specifically developed a rationale that integrates the ecological and anthropogenic dimensions of marginality to determine the regional and local scale mechanisms shaping the probability of persistence (or extinction) of oak rear edge populations. Our main hypothesis is that these stands will show low resilience to extreme droughts along a small-scale gradient. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their present geographical rear edge; (*ii*) conducted and explored a review of historical documents to reconstruct forest management history and to infer how it impacted tree-growth in time; (*iii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iv*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter, we characterized variability in the forest response to drought within the current geographical rear-edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.

@@

The forests of this species reach their southernmost European limit in Andalusian mountains such as Sierra Nevada (37°N, 3°W), a high-mountain range with elevations of up to 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought and increasing aridity with decreasing altitude, and marked variability according to elevation and aspect. Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olalde2002]. Eight Pyrenean oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys. Today, *Quercus pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005]. They are the richest forest formation in vascular plant species of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008ABG]. They also harbor high levels of intraspecific genetic diversity [@Valbuena2013]. These relict forests have undergone intensive human use throughout history [@CamachoOlmedo2002]. Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013; @GeaIzquierdoetal2017RiskyFuture].

@@

### Field sampling and dendroecological methods to assess individual tree resilience
Trees were sampled during the autumn of 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect (around 1400 m); and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1). For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit for the species (CA-Low) and the maximum altitude currently reached by trees (CA-High), respectively, in the site sampled. Despite the proximity of these two elevations (less than a 200-m difference) the stands differ markedly in their structure and characteristics (Table 1). The three sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015]. At each site, between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly sampled. Two cores of 5 mm in diameter were taken from each tree at breast height (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species, and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot with a 10-m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details].

@@

The growth trends were analyzed at different time scales. To study the growth response to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. These were calculated from ratios between raw growth measurements and individual cubic splines with a 50% frequency cutoff at 30 years [@Fritts1976]. Tree-ring width series were standardized and detrended using dplR [@Bunn2010]. Mean residual site chronologies were established by computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976]. The statistical quality of each chronology was checked via the expressed population signal (EPS). A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.

The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$). In theory, BAI represents a more accurate indicator of growth than ring width, because it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile growth [@Biondi2008]. Raw ring widths and measured DBH were used to compute BAI [@Piovesan2008] with the following equation: $\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$ where $r$ is the radius of the tree and $t$ is the year of tree-ring formation. For each individual tree, a mean BAI series was calculated. Then, mean site BAI chronologies were determined by averaging individual tree BAI time series.

### Disturbance analyses and land-use history review 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (suppressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought-induced neighbor mortality) in the past. Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median [@Rubino2004].
```{r, eval=FALSE, echo=FALSE}
# Medians were used since they are a more robust estimator of central tendency than means [@Rubino2004].
```
Site-disturbance chronologies were constructed by annually averaging the individual disturbance series. To separate growth peaks caused by disturbance events and expressing stand-wise disturbances from those caused by climate, we considered a threshold of 50% of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014]. In addition, the history of the forest and management of our sampling sites was inferred from a detailed analysis of historical land-use changes. For this, existing historical documents were exhaustively reviewed to compile information on socio-economical activities affecting the forests being studied (Table S4). We exhaustively reviewed existing documentary sources: historical documents and maps; detailed mining reports; official information on recent wildfires events and forest-management practices; livestock farming; traditional irrigation channels; and studies concerning the socioeconomic dynamics of forests on Sierra Nevada at different spatio-temporal scales (see Table S4 for references).

@@

### Statistical analysis

@@

The severe drought events since 1901 were identified using SPEI-12. They were characterized in terms of duration, severity and intensity [see @Spinoni2015]. In a first step, the impact of drought in greenness and growth was explored using the $EVI_{sa}$ and the mean RWI site chronologies (Figure 2). Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim [@Zang2015]. The non-climatic disturbance impacts on tree-growth were evaluated using site disturbance chronologies (buit using growth changes on tree growth).

Responses of vegetation to disturbances were explored in the short- and the long-term using resilience metrics (resilience, resistance and recovery) and temporal trends respectively for both EVI and BAI (Figure 2). Resilience metrics of BAI were computed for the eight most severe drought events since 1950 (including 2005 and 2012), and their relationship with drought severity were explored. Resilience metrics of EVI were computed only for 2005 and 2012 drought events. Temporal trends of $EVI_{mean}$ (pixel scale) and BAI (mean BAI site chronologies) were examined using non-parametric (Mann-Kendall) and parametric test (Pearson) respectively.

For each of the three resilience indices studied, we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the oak populations studied (northern and southern exposures). These tests were used because original and log-transformed data did not follow the assumptions of normality or homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used because they were close to the mean value in all cases [@Wilcox2012]. When the robust ANOVA test was run, data were bootstrapped 3000 times and trimmed automatically to control the potential influence of outliers. *Post-hoc* differences were assessed pairwise using a similar bootstrap test. All the robust ANOVA and *post-hoc* tests were carried out using the WRS2 R package [@Mair2017]. The level of significance was set to 0.05 and adjusted for multiple comparisons.
